How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2021-07-20 00-author.Rmd
Add files via upload
# After copy edits

```{r fig.align='right', echo=FALSE, include=identical(knitr:::pandoc_to(), 'html'), fig.link='http://john.measey.com/About-Me'}
knitr::include_graphics('figures/john.jpg', dpi = NA)
```

# Do you have something to contribute? {- #contribute}

This book is written in bookdown [@xie2016bookdown] specifically to make it a 'live project' that will be open to anyone who wants to contribute, improve, or use as the basis for your own book. One of the amazing potentials for bookdown books is that all the files for this book are hosted on [Github](https://github.com/johnmeasey/How-to-write-a-PhD-in-Biological-Sciences/tree/main). You have the opportunity to fork this repository and write your own version for a different discipline, a different language or for a different region of the world. It would be impossible to cover every aspect of writing a PhD in Biological Sciences, and it may be that I have missed out ones that are very important to you. Equally, parts of what is currently written will become obsolete very quickly as new initiatives are begun, and old problems are resolved. For this reason, this guide needs to be a 'living document', and anyone who wants to provide feedback or contribute new sections is more than welcome. Please feel free to open an issue, or make a Pull Request if you spot a typo.

>>>>

```{r fig.align='right', out.width = '15%', echo=FALSE, include=identical(knitr:::pandoc_to(), 'html'), fig.link='http://john.measey.com/About-Me'}
knitr::include_graphics('figures/john.jpg', dpi = NA)
```

## Do you have something to contribute? {- #contribute}

This book is written in bookdown [@xie2016bookdown] specifically to make it a 'live project' that will be open to anyone who wants to contribute, improve, or use as the basis for your own book. The easiest way for readers to contribute content directly is through a [GitHub pull requests](https://help.github.com/articles/about-pull-requests/). At the repository for this book, you will find Rmd files for each chapter, and as a GitHub user, you can simply edit the Rmd file and submit the changes. If I am happy with the changes proposed, I will merge your content with that of the book and add your name to the [Acknowledgements](#acknowledge).

One of the amazing potentials for [bookdown books](https://bookdown.org/) is that all the files for this book are hosted in a repository on [Github](https://github.com/johnmeasey/How-to-write-a-PhD-in-Biological-Sciences/tree/main). You have the opportunity to fork this repository and write your own version for a different discipline, a different language or for a different region of the world. It is also my hope that this guide can grow to become a community of practice for those conducting PhDs in Biological Sciences. It will not be possible to cover every aspect of writing a PhD in Biological Sciences, but it may be that I have missed out ones that are very important to you. Equally, parts of what is currently written will become obsolete as new initiatives begin, and old problems are resolved. For this reason, this guide needs to be a 'living document', and anyone who wants to provide feedback or contribute new sections is more than welcome. Please feel free to open an issue, or make a Pull Request if you spot a typo.

## Creative Commons License {-}
How to write a PhD in Biological Sciences: a guide for the uninitiated is copyright to John Measey and distributed under a Creative Commons BY-NC-ND 4.0 International License. You may freely share this content for non-commercial purposes, with a source credit to [www.HowToWriteAPhD.org](www.HowToWriteAPhD.org).

## Disclaimer {-}

Although I have tried to make the information is this book as accurate as possible, it is provided without any warranty. The author and publisher have neither liability nor responsibility to any person or entity related to any loss or damages arising from the information contained in this book.

```{r iris}
knitr::kable(
  head(iris), caption = 'We must have the iris data and some R code, otherwise it would not seem right',
  booktabs = TRUE
)
```

### 2021-07-20 1.01-beforeyoustart.Rmd
Add files via upload
# After copy edits

---
output:
  html_document: default
  pdf_document: default
---
```{r iris}
knitr::kable(
  head(iris), caption = 'We must have the iris data and some R code, otherwise it would not seem right',
  booktabs = TRUE
)
```

@@

 - The median time to completion in many countries worldwide is much longer than it should be [e.g. 7.5 years in the USA @geven2018how].
 - Completion rates are shockingly low [as low as 50% in the USA @litalien2015improving].

This has spurred many responsible for postgraduate students to try to ensure that they receive more support for their studies. But there is also the need for students to comprehend right from the start that they are undertaken a major commitment that will likely impact many different parts of their lives.

In this first chapter of the book I intend to make you familiar with what a PhD is, how to approach thinking about it, and what to look for in an institution and advisor. You may already have some or all of these things fixed, but there is still some use in going over these basic points. Try to keep an open mind. Doing a PhD will take several years of your life, and may well define your future career prospects. Therefore, to say that this is a big decision is an understatement. It is not a substitute for getting a job or a way of remaining a student for longer. If after reading this section you are in any doubt about whether or not a PhD is right for you, then I would suggest that you do need to think about it further. Don't take my word for it, go and talk to people who have done a PhD (whether or not they now have a job), or those who are doing a PhD now. Especially if their experience is in your focal research area, they are likely to know a lot more about it.

@@

## Is a PhD more than a thesis?
Yes. The thesis is really only the final evidence of all of the work and experience that you undertake while doing your PhD. The time that you spend doing your PhD will likely include participating in conferences, giving talks, participating in journal clubs, seminars, workshops and working groups. In addition to all of the writing skills, on which this book focuses, you will develop skills in techniques that relate to your particular area, and learn to communicate this to different types of audience. You will also learn and develop your skills in presentation, analysis and treatment of data. It is an intense learning experience. It should also be a fun and positive experience.

@@

## Criteria used to judge PhD studies
In order to provide you with some general idea of what is required, I list below the (edited) questions asked of examiners for a PhD thesis by my university:

- Has the motivation for the objectives of the study been formulated satisfactorily?
- Do the research results constitute a meaningful contribution to the knowledge of and insight into the relevant field of study?
- Does the dissertation distinguish clearly between own, new contributions to, and known results in the relevant field of study?
- Is the candidate capable of evaluating the scientific meaning of his / her results and of placing this in context within existing knowledge in the field of study?
- Does the candidate show signs of independent, critical thinking or other signs of originality?
- Does the candidate show that they are sufficiently capable of doing independent research?
- Does the dissertation show that the candidate is sufficiently familiar with the relevant research techniques and methods?
- Does the dissertation show conversance with, and a critical attitude towards the pertinent literature?
- Is the material presented in a clear, systematic and logical manner?
- Is the linguistic, stylistic and technical editing of the dissertation acceptable?
- Are the research results acceptable for publication?

If you compare these PhD requirements with those of an MSc, the key difference is that for a PhD you are required to produce a "meaningful contribution to the knowledge of and insight into the relevant field of study", and show that you are "...capable of doing independent research". Thus, a PhD degree prepares you to be an independent researcher, and so you are expected to have a critical attitude toward other research.

@@

- First is the clear ability to conceptualise and carry out a body of scientific work that is novel. Here I mean novel in that it has not simply been a repeat of some previous work on a different taxon or system. This is easier to judge from an oral defense, over reading the thesis. 
- Second is the clear knowledge of how the work that is conducted fits into the bigger picture. That is that the work demonstrates that the person is a scientist at a broader level than the narrow focus of their research.
- Third is whether the student has been able to bring the study together in a way that shows that it is more than the sum of its parts. In essence, together the 5 data chapters should provide a bigger overview of a broader topic, and show the direction of future work.

@@

## What will you do after your PhD?
It sounds silly to think about what comes after your PhD just as you are just starting, but the reality is that you always need to be planning ahead in order to make the most from what you are doing now. During your PhD you will have lots of opportunities in many different areas of academia, teaching and practically applied areas. Knowing what you want to do when you finish will help guide your decisions so that you can maximise opportunities as they arise. Both inside and outside of academia, it is likely to be the contacts that you make that will help in your next career move.

@@

At the same time, governments the world over, and especially developing countries, are providing bursaries for people to do PhDs. Why? The number of PhDs in a country is considered  (by the [World Bank](https://datacatalog.worldbank.org/dataset/world-development-indicators)) to be one of many measures of the development of that country. As the percentage gets larger, so countries can convince the world bank that they are developing (and then get granted its favours, see Figure \@ref(fig:PhDMap)). This is all well and good, but strategically, governments should consider why they want these researchers in their populations, and more particularly for what kinds of jobs. It seems somewhat unfair to produce lots of young PhD holders that have no careers to join, or prospects to use their PhD just so that your government can tick a check box on their road to development, and so borrow more money from the World Bank. If you are from a developing country, you cannot assume that just because you are given a bursary by your government it follows that they will give you a job or a career. This will likely always be your own task.

(ref:PhDMap) **Where in the world are there most PhDs?** In this map of the world, countries shaded are those that have data at the declaring the proportion of PhD (or equivalent) in their population (Max percentage between 2010 and 2020 (Data from the World Bank). This ranges from 2.97% in Switzerland (dark red) to 0.02% in DR Congo (light yellow). Notably there is no data held in these years for many countries in Asia, Africa and South America.

@@

It is important to realise that many academics are looking for good students to do PhD work. This is because they need someone to do research within their groups, and students or postdocs do most of this work [see @chamberlain2016ten]. For this reason, an academic you know while you're doing your MSc studies may start to try to persuade you to do a PhD. This could be fantastic for you, it could be just what you're looking for. But the reality is that you will never do a PhD for somebody else, you are always going to be doing it for yourself. if the only reason you are doing a PhD is because someone else told you that you should then you are likely to have a really hard time. By the time you graduate, you may have a PhD but without any idea of what you're going to do with it.

@@

You should be aware by now that there are lots of people in academia who are ‘institutionalised’. By this I mean that they have never had a job (or a life) outside of academia. This is not something that I can recommend for anyone. Such people are quite peculiar in many ways, and they tend to operate outside of the norms and standards of the rest of society. In particular, they make academia a more difficult place to work, because they tend to accentuate and maintain a lot of the silliness within these institutions. If you are worried about becoming such a person, then I think it would be advisable to take some time outside of academia. For example, if you have never worked but are thinking of doing a PhD, consider taking a year (or two) to do a job before starting a PhD. If it is in a related field it will likely help you greatly. I suggest that it will certainly help keep your feet on the ground, and in touch with some level of reality or normality, during your PhD. You’ll also have some experience and connections to go back to once you’ve finished.

## Don't rely on becoming an academic
Academia is getting more competitive. Whereas once it was almost certain that you would be employed as an academic after finishing your PhD, this is no longer the case. Indeed, the number of people who get PhDs is greatly increasing the world over, and most are in direct competition for a decreasing amount of academic posts. In 2019, 39% of respondents to a survey said that the most important reason they decided to enrol in a PhD programme was to pursue an academic career [@woolston2019phds]. Further, 56% wanted academia as their first choice for a career. Yet in Belgium, for example, the numbers of PhD graduates entering the market nearly doubled (in the first 15 years of the 21st century) while academic positions remained static [@levecque2017work]. Universities appear to be exploiting this excess of qualified people to hire them on contracts, reducing the number of full time positions. Essentially, this means that in most places your time as a PhD student, postdoc and junior lecturer will keep you outside the employers core staff. Joining those on the inside (with well funded posts) becomes increasingly more difficult. Some have suggested that the world of academia is becoming more like a gang producing drugs [@afonso2014academia].

## Doing a PhD is really hard work
It’s not a MSc with an extra year. It really is much harder and tougher on you mentally and physically. The chances that you’ll end up jaded and malcontent with the entire academic system are very high (maybe even obligatory). You are probably not going to be very happy while doing it (there are some exceptions of course), and it will likely have negative impacts on all the good things that you currently have in your life (think family, friends and relationships, as well as free time and fun).

## Your mental health will likely suffer
Doing a PhD will be the cause of stress not only in your working life, but will also impact other areas of your life. Being aware of this at the outset will allow you to alert close members of your non-working life (think partner, family and friends) that they may well need to act as a support network. It is worth tracking your mental health during your PhD to check that you are not getting into difficulties.

@@

The country where you study a PhD does not determine the quality of institution (see Figure \@ref(fig:THEreseasrchcountry). Rather look at the merit of the particular institution that you are interested in. Surveys like that from the [Times Higher Education Supplement](https://www.timeshighereducation.com/world-university-rankings/2021/world-ranking) can help give you an overview.

(ref:THEreseasrchcountry) **It’s not the country that determines the quality of your university.** In this plot you see the Times Higher Education (THE) World Ranking (2021) score for Research plotted by country (with ≥ 10 universities). Note that most universities in most countries have mean THE Research scores in the lowest quartile. Almost all countries that have institutions in the top quartile, have most of the institutions in that country in the bottom quartile. The point here is that you should look at the quality of the institution that you want to attend, rather than assume that all universities from certain countries have high or low research profiles.

@@

## What to look for in an advisor
Throughout this book I will call the person who is the academic in charge of your PhD studies your ‘advisor’. I will not call them your supervisor. The reason for this is because I think it is much better for you to think of this person as giving you advice and not the person who tells you what to do. This comes back to the idea that the PhD is yours. It is not your advisor’s PhD. They will not write it. They will not do all the work that's required for it. That will be yours alone.

Just as I've attempted to do in this book, that person will give you advice. It will be up to you whether or not you follow that advice. Many of my students have not followed my advice. For some this has been the right thing to do. For others it has not. No one can know what the best thing to do is in different situations, because they will almost always be unique. You will be the expert on your project. That means that as the person closest to your project, you are the person who is best placed to make a decision. But your advisor can, and should, help you to make those decisions.

@@

One of the best ways of picking a good advisor is by talking to current or past students. You should ask them how happy they are with their advisor and whether they are receiving the support that they need to do their projects. A good advisor should have an open door so that you can talk to them whenever you need to. Thus, once you have talked to the students in their group, you should also make an appointment to talk to them about your project. It might not be one meeting, there may be several. The more contact you have with them, the more likely you are to be sure that this is the person whose help and advice that you want when conducting research over the next few years.

@@

At the end of the day there is no such thing as the perfect advisor. You will likely have to make a trade-off on the above points with someone who has a project that they are willing to advise you on, and your chosen area of research.

## What makes a good project?
After having made lots of statements about how a project should be your own project the following advice might seem a bit strange. Clearly the work that you're going to be doing will need to fit into the group of the advisor that you've chosen.

The most obvious point to make here is that a project should be something that you want to do and that you're prepared to dedicate yourself to for ~3 years.

@@

## Funding your PhD studies
Funding may be one of the biggest obstacles that you come across when wanting to do a PhD. It's important that you consider the funding of your PhD in two separate ways:

### The funding that you need to live 
This money is typically called a bursary or stipend depending on where you are in the world. It seems to me that many countries offer bursaries that are not actually enough to live on (e.g. buy food, rent a house etc.). If you're offered a PhD bursary, especially if it's in another country, then do check on the living cost at that location and whether or not the bursary that you are being offered will be sufficient money in order for you to survive. Also, be very careful to check how long the bursary is for. For example, at my university a PhD bursary is given only for 3 years. After this three-year funding period, you will be expected to have submitted your PhD for examination. If you need further funding to live while your PhD is being examined, and before you graduate then you will need to get a job. There is no more money!

It's a really good idea to talk to people who have already done a PhD to find out more about the different available bursary opportunities, and how people manage with and without them. There is no one single model, and the more people you talk to the more likely you are to discover one that might suit you. Of course conferring with other people can include reading the accounts of many different people who share their experiences on social media. There are blogs and Twitter accounts dedicated to this stuff. Try and be as open-minded as you can.

### The cost associated with your studies. 
The money associated with the project that you were doing also needs to come from somewhere. Depending on where in the world you can do these studies they will be different expectations of where this money comes from. This is a very important topic for you and your potential advisor to discuss. During the proposal phase of your PhD you'll be expected to put together a budget to determine the total costs of your studies. In reality you or your advisor should have a ballpark figure of how much this Is going to cost and have some idea about where the money is going to come from.

@@

It is possible that your potential advisor will tell you that they don't have any funding for a project that you may want to do. This is both a curse and an opportunity. There are many opportunities to raise money to fund PhD projects. Raising the funding yourself is one way of making sure it really remains your project. But the availability of this kind of funding does vary from country to country and you may want to embark on the fundraising side of your PhD before you register. Especially if you have a limited amount of time once you have registered in order to complete your PhD studies. 
Some projects are much cheaper than others. For example, if your intention is to sequence genomes of every animal or plant that you work with, the finance required will likely be far greater than your bursary. If however your intention is to do a PhD that models distributions of populations of animals and plants, then the entire costs may only run to a computer remote access to a computing cluster and data storage.

>>>>

---
output:
  html_document: default
  pdf_document: default
---

@@

 - The time to completion in many countries worldwide is much longer than it should be. For example, 7.5 years is the median time to completion in the USA [@geven2018how], Europe universities have largely adopted the Salzburg Principles that doctoral programmes should be three to four years for full-time candidates  [@christensen2005bologna]. Following this, an average is recorded as being between 3.5 and 4.5 years, with most institutions having reduced this time over the last 10 years [@hasgall2019doctoral].
 - Completion rates are shockingly low. As bad as 50% in the USA [@litalien2015improving], 66% (within 6 years) in Europe [@hasgall2019doctoral], but generally better in the sciences [@park2005new].

This has spurred many responsible for postgraduate students to try to ensure that they receive more support for their studies. But there is also the need for students to comprehend right from the start that they are undertaking a major commitment that will likely impact many different parts of their lives.

In this first chapter of the book, I intend to make you familiar with what a PhD is, how to approach thinking about it, and what to look for in an institution and advisor. You may already have some or all of these things fixed, but there is still some use in going over these basic points. Try to keep an open mind. Doing a PhD will take several years of your life, and may well define your future career prospects. Therefore, to say that this is a big decision is an understatement. It is not a substitute for getting a job or a way of remaining a student for longer. If after reading this section you are in any doubt about whether or not a PhD is right for you, then I would suggest that you do need to think about it further. Don't take my word for it, go and talk to people who have done a PhD (whether or not they now have a job), or those who are doing a PhD now. Especially if their experience is in your focal research area, they are likely to know a lot more about it.

@@

## Is a PhD more than a thesis?
Yes. The thesis is really only the final evidence of all of the work and experience that you undertake while doing your PhD. The time that you spend doing your PhD will likely include participating in conferences, giving talks, participating in journal clubs, seminars, workshops and working groups. In addition to all of the writing skills, on which this book focuses, you will develop skills in techniques that relate to your particular subject area, and learn to communicate this to different types of audience. You will also learn and develop your skills in presentation, analysis and treatment of data. It is an intense learning experience. It should also be a fun and positive experience.

@@

## Criteria used to judge PhD studies {#criteria}
In order to provide you with some general idea of what is required, I list below the (edited) questions asked of examiners for a PhD thesis by my university (you should ask for the criteria at your own institution):

- Has the motivation for the objectives of the study been formulated satisfactorily?
- Do the research results constitute a meaningful contribution to the knowledge of and insight into the relevant field of study?
- Does the dissertation distinguish clearly between own, new contributions to, and known results in the relevant field of study?
- Is the candidate capable of evaluating the scientific meaning of their results and of placing this in context within existing knowledge in the field of study?
- Does the candidate show signs of independent, critical thinking or other signs of originality?
- Does the candidate show that they are sufficiently capable of doing independent research?
- Does the dissertation show that the candidate is sufficiently familiar with the relevant research techniques and methods?
- Does the dissertation show conversance with, and a critical attitude towards the pertinent literature?
- Is the material presented in a clear, systematic and logical manner?
- Is the linguistic, stylistic and technical editing of the dissertation acceptable?
- Are the research results acceptable for publication?

If you compare these PhD requirements with those of an MSc, the key difference is that for a PhD you are required to produce a "meaningful contribution to the knowledge of and insight into the relevant field of study", and show that you are "...capable of doing independent research". Thus, a PhD degree prepares you to be an independent researcher, and so you are expected to have a critical attitude towards other research.

@@

- First is the clear ability to conceptualise and carry out a body of scientific work that is novel. Here I mean novel in that it has not simply been a repeat of some previous work on a different taxon or system. This is easier to judge from an oral defense, over reading the thesis. 
- Second is the clear knowledge of how the work that is conducted fits into the bigger picture. That is, that the work demonstrates that the person is a scientist at a broader level than the narrow focus of their research.
- Third is whether the student has been able to bring the study together in a way that shows that it is more than the sum of its parts. In essence, together the 5 data chapters should provide a bigger overview of a broader topic, and show the direction of future work.

@@

## What will you do after your PhD?
It sounds silly to think about what comes after your PhD just as you are just starting, but the reality is that you always need to be planning ahead in order to make the most of what you are doing now. During your PhD you will have lots of opportunities in many different areas of academia, teaching and practically applied areas. Knowing what you want to do when you finish will help guide your decisions so that you can maximise opportunities as they arise. Both inside and outside of academia, it is likely to be the contacts that you make that will help in your next career move.

@@

At the same time, governments the world over, and especially developing countries, are providing bursaries for people to do PhDs. Why? The number of PhDs in a country is considered  (by the [World Bank](https://datacatalog.worldbank.org/dataset/world-development-indicators)) to be one of many measures of the development of that country. As the percentage gets larger, so countries can convince the World Bank that they are developing (and then get granted its favours, see Figure \@ref(fig:PhDMap)). This is all well and good, but strategically, governments should consider why they want these researchers in their populations, and more particularly for what kinds of jobs. It seems somewhat unfair to produce lots of young PhD holders that have no careers to join, or prospects to use their PhD just so that your government can tick a check box on their road to development, and so borrow more money from the World Bank. If you are from a developing country, you cannot assume that just because you are given a bursary by your government it follows that they will give you a job or a career. This will likely always be your own task.

(ref:PhDMap) **Where in the world are there most PhDs?** In this map of the world, countries shaded are those that have data declaring the proportion of PhD (or equivalent) in their population (max percentage between 2010 and 2020 (Data from the World Bank). This ranges from 2.97% in Switzerland (dark red) to 0.02% in DR Congo (light yellow). Notably, there is no data held in these years for many countries in Asia, Africa and South America.

@@

It is important to realise that many academics are looking for good students to do PhD work. This is because they need someone to do research within their groups, and students or postdocs do most of this work [see @chamberlain2016ten]. For this reason, an academic you know while you're doing your MSc studies may start to try to persuade you to do a PhD. This could be fantastic for you, it could be just what you're looking for. But the reality is that you will never do a PhD for somebody else, you are always going to be doing it for yourself. If the only reason you are doing a PhD is because someone else told you that you should then you are likely to have a really hard time. By the time you graduate, you may have a PhD but without any idea of what you're going to do with it.

@@

You should be aware by now that there are lots of people in academia who are ‘institutionalised’. By this, I mean that they have never had a job (or a life) outside of academia. This is not something that I can recommend for anyone. Such people are quite peculiar in many ways, and they tend to operate outside of the norms and standards of the rest of society. In particular, they make academia a more difficult place to work, because they tend to accentuate and maintain a lot of the silliness within these institutions. If you are worried about becoming such a person, then I think it would be advisable to take some time outside of academia. For example, if you have never worked but are thinking of doing a PhD, consider taking a year (or two) to do a job before starting a PhD. If it is in a related field it will likely help you greatly. I suggest that it will certainly help keep your feet on the ground, and in touch with some level of reality or normality, during your PhD. You’ll also have some experience and connections to go back to once you’ve finished.

## Don't rely on becoming an academic
Academia is getting more competitive. Whereas once it was almost certain that you would be employed as an academic after finishing your PhD, this is no longer the case. Indeed, the number of people who get PhDs is greatly increasing the world over, and most are in direct competition for a decreasing amount of academic posts. In 2019, 39% of respondents to a survey said that the most important reason they decided to enrol in a PhD programme was to pursue an academic career [@woolston2019phds]. Further, 56% wanted academia as their first choice for a career. Yet in Belgium, for example, the numbers of PhD graduates entering the market nearly doubled (in the first 15 years of the 21st century) while academic positions remained static [@levecque2017work]. Universities appear to be exploiting this excess of qualified people to hire them on contracts, reducing the number of full-time positions. Essentially, this means that in most places your time as a PhD student, postdoc and junior lecturer will keep you outside the employers core staff. Joining those on the inside (with well-funded posts) becomes increasingly more difficult. Some have suggested that the world of academia is becoming more like a gang producing drugs, with those tenured academics earning good, secure salaries at the top of the pyramid, while increasing numbers of insecure poorly paid doctoral students and post-docs do most of the actual work [@afonso2014academia].

## Doing a PhD is really hard work
It’s not an MSc with an extra year. It really is much harder and tougher on you [mentally and physically](#healthy). The chances that you’ll end up jaded and malcontent with the entire academic system are very high (maybe even obligatory). At times, you are probably not going to be very happy while doing it (there are some exceptions of course), and it will likely have negative impacts on all the good things that you currently have in your life (think family, friends and relationships, as well as free time and fun). On the other hand, it can simultaneously bring joy and start you off on a lifetime trajectory.

## Your mental health will likely suffer
Doing a PhD will be the cause of stress not only in your working life, but will also impact other areas of your life. Being aware of this at the outset will allow you to alert close members of your non-working life (think partner, family and friends) that they may well need to act as a support network. It is worth tracking your mental health during your PhD to check that you are not getting into difficulties, and I show you later [how to do this](#healthy).

@@

The country where you study a PhD does not determine the quality of institution (see Figure \@ref(fig:THEreseasrchcountry)). Rather look at the merit of the particular institution that you are interested in. Surveys like that from the [Times Higher Education Supplement](https://www.timeshighereducation.com/world-university-rankings/2021/world-ranking) can help give you an overview.

(ref:THEreseasrchcountry) **It’s not the country that determines the quality of your university.** In this plot, you see the Times Higher Education (THE) World Ranking (2021) score for Research plotted by country (with ≥ 10 universities). Note that most universities in most countries have mean THE Research scores in the lowest quartile. Almost all countries that have institutions in the top quartile, have most of the institutions in that country in the bottom quartile. The point here is that you should look at the quality of the institution that you want to attend, rather than assume that all universities from certain countries have high or low research profiles.

@@

## What to look for in an advisor
Throughout this book, I will call the person who is the academic in charge of your PhD studies your ‘advisor’. I will not call them your supervisor. The reason for this is because I think it is much better for you to think of this person as giving you advice and not the person who tells you what to do. This comes back to the idea that the PhD is yours. It is not your advisor’s PhD. They will not write it. They will not do all the work that's required for it. That will be yours alone.

Just as I've attempted to do in this book, that person will give you advice. It will be up to you whether or not you follow that advice. Many of my students have not followed my advice. For some, this has been the right thing to do. For others, it has not. No one can know what the best thing to do is in different situations, because they will almost always be unique. You will be the expert on your project. That means that as the person closest to your project, you are the person who is best placed to make a decision. But your advisor can, and should, help you to make those decisions.

@@

One of the best ways of picking a good advisor is by talking to past students, or current students that have been in the lab a while. You should ask them how happy they are with their advisor and whether they are receiving the support that they need to do their projects. A good advisor should have an open door so that you can talk to them whenever you need to. Thus, once you have talked to the students in their group, you should also make an appointment to talk to them about your project. It might not be one meeting, there may be several. The more contact you have with them, the more likely you are to be sure that this is the person whose help and advice that you want when conducting research over the next few years.

@@

At the end of the day, there is no such thing as the perfect advisor. You will likely have to make a trade-off on the above points with someone who has a project that they are willing to advise you on, and your chosen area of research.

## What makes a good project?
The most obvious point to make here is that a project should be something that you want to do and that you're prepared to dedicate yourself to for ~3 years.

@@

## Funding your PhD studies
Funding may be one of the biggest obstacles that you come across when wanting to do a PhD. It's important that you consider the funding of your PhD in two separate ways.

### The funding that you need to live 
This money is typically called a bursary or stipend depending on where you are in the world. It seems to me that many countries offer bursaries that are not actually enough to live on (e.g. buy food, rent a house, etc.). If you're offered a PhD bursary, especially if it's in another country, then do check on the living cost at that location and whether or not the bursary that you are being offered will be sufficient money in order for you to survive. Also, be very careful to check how long the bursary is for. For example, at my university a PhD bursary is given only for 3 years. After this three-year funding period, you will be expected to have submitted your PhD for examination. If you need further funding to live while your PhD is being examined, and before you graduate then you will need to get a job. There is no more money!

It's a really good idea to talk to people who have already done a PhD to find out more about the different available bursary opportunities, and how people manage with and without them. There is no one single model, and the more people you talk to the more likely you are to discover one that might suit you. Of course, conferring with other people can include reading the accounts of many different people who share their experiences on social media. There are blogs and Twitter accounts dedicated to this stuff. Try and be as open-minded as you can.

### The cost associated with your studies. 
The money associated with the project that you are doing also needs to come from somewhere. Depending on where in the world you can do these studies there will be different expectations of where this money comes from. This is a very important topic for you and your potential advisor to discuss. During the proposal phase of your PhD you'll be expected to put together a budget to determine the total costs of your studies. You or your advisor should have a ballpark figure of how much this is going to cost and have some idea about where the money is going to come from.

@@

It is possible that your potential advisor will tell you that they don't have any funding for a project that you may want to do. This is both a curse and an opportunity. There are many opportunities to raise money to fund PhD projects. Raising the funding yourself is one way of making sure it really remains your project. But the availability of this kind of funding does vary from country to country and you may want to embark on the fundraising side of your PhD before you register. Especially if you have a limited amount of time once you have registered in order to complete your PhD studies. 
Some projects are much cheaper than others. For example, if your intention is to sequence genomes of every animal or plant that you work with, the finance required will likely be far greater than your bursary. If however your intention is to do a PhD that models distributions of populations of animals and plants, then the entire costs for your study may only add up to a computer, remote access to a computing cluster and data storage.

### 2021-07-20 1.02-useR.Rmd
Add files via upload
# After copy edits

# Reconciling yourself to doing things you've been avoiding {#useR}
Some 30 odd years ago as an undergraduate I was told that I would have to use computers in one of my 2nd year courses. The thought filled me with dread and I remember putting off going into the computer room for as long as I could. It was a steep learning curve, and I struggled. Each time the cursor got to the end of the screen I hit the carriage return button. And each machine had an odd white device attached called a mouse, and I had no idea what that did. All this seems quite preposterous now, but imagine embarking on a PhD in biological sciences without reconciling yourself to using a PC. This is practically where we are now with R [@rcoreteam2021r], and indeed a wealth of other alternative programming platforms.

@@

If you are not already aware of the prominence that R now has in conducting statistical analyses in biological sciences, then you’ve missed out on a great deal. The real point here is not that you can only use R, but that R represents a platform that can be easily reproduced (because it’s free unlike many statistical software packages), and by saving and making the code available, you are making your data analyses reproducible. We will see elsewhere the importance of reproducibility in sciences [@baker20161500], and the critical role this has in transparency and therefore best practice.

@@

If you already know all of the above, then you won’t face any major barriers, and you will fit into the new and more transparent world of science with great ease. If the thought of using R fills you with dread, in the same way that using a computer did some 30 years ago, then you should reconcile yourself now before you go any further. Learning R will be a steep learning curve, but it will be made easier by the large amount of excellent video tutorials and online courses that are available.

## And learning lots of other stuff...
The same principle that I’ve written about above can be extended to all sorts of other aspects of your PhD project. You will have to learn new skills, and some of these will be things that you have previously resisted having to do. Allow yourself to be open to all aspects of learning, and all skills. The act of learning opens new pathways in your brain, and using this ‘muscle’ will facilitate future learning exercises.

>>>>

# Reconciling yourself to doing things you've been avoiding {#useR}

Some 30 odd years ago, as an undergraduate, I was told that I would have to use computers in one of my 2nd year courses. The thought filled me with dread and I remember putting off going into the computer room for as long as I could. It was a steep learning curve, and I struggled. As my only previous experience had been with a typewriter, each time the cursor got to the end of the screen I hit the 'carriage-return' button. And each machine had an odd white device attached called a mouse, and I had no idea what that did. All this seems quite preposterous now, but imagine embarking on a PhD in biological sciences without reconciling yourself to using a PC. This is practically where we are now with R (@rcoreteam2021r; Figure \@ref(fig:Rlogo)), and a wealth of other emerging and alternative programming platforms.

@@

If you are not already aware of the prominence that R now has in conducting statistical analyses in biological sciences, then you’ve missed out on a great deal. The real point here is not that you can use R, but that R represents a platform on which analyses can be easily repeated, and by saving and making the code available together with your data, you are making your data analyses reproducible. We will see elsewhere the importance of reproducibility in sciences [@baker20161500], and the critical role this has in transparency and, therefore, best practice.

@@

If you already know all of the above, then you won’t face any major barriers, and you will fit into the new and more transparent world of science with great ease. If the thought of using R fills you with dread, in the same way that using a computer did for me some 30 years ago, then you should reconcile yourself now before you go any further. Learning R will be a steep learning curve, but it will be made easier by the large amount of excellent video tutorials and online courses that are available.

## And learning lots of other stuff
The same principle that I’ve written about above can be extended to all sorts of other aspects of your PhD project. You will have to learn new skills, and some of these will be things that you have previously resisted having to do. Allow yourself to be open to all aspects of learning, and all skills. The act of learning opens new pathways in your brain, and using this ‘muscle’ will facilitate future learning exercises.
